---
title: "REKN altitude mapping"
author: "Gen Perkins"
date: "2025-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#library("rnaturalearth")
#library("rnaturalearthdata")
library(lubridate)
library(sf)
library(stringr)
library(readr)
library(dplyr)
library(ggplot2)
library(ggspatial)
library(leaflet)
library(fs)


```

## Altitude Mapping for Red Knots. 

While the use of high accuracy GPS location data has enabled better understanding of speed and distance of long distance migratory birds, such as Red Knots. 

Despite these developments, an understanding of bird altitude movements remains less understood.

In this project we used Red Knot GPS data, along with modeled wind speed and directions to assemble data on both bird movement (speed, direction, bearing) and ambient wind conditions. We specifically were interested in offshore movements, and estimated wind speed and direction using the RSS Cross-Calibrated Multi-Platform (CCMP) 6-hourly ocean vector wind analysis on 0.25 deg grid, Version 3.0 
(see:<https://doi.org/10.56236/RSS-uv6h30>).


```{r load data, include=FALSE}

bb <- st_read(path("02_outputs", "birds_final_proc","birdwind_all_20250426.gpkg"), quiet = TRUE) 

# format the data 

bb <- bb |>
  select(
    id, filename, datetime, shore_status,Fix, Alt.m., bearing, speed_kmh, geoid_height, ortho_height,
    ws, windangle, windangle_azmith, geom
  ) |>
  dplyr::mutate(
    bid = stringr::str_split_i(filename, "_", 4),
    bproj = stringr::str_split_i(filename, "-", 2)
  ) |>
  select(-c(filename))



# format colours 
bb <- bb |> 
  mutate(ortho_height_cat = case_when(
    ortho_height < 50 ~ "0 - 50",
    ortho_height >= 50 & ortho_height < 100 ~ "50 - 100",
    ortho_height >= 100 & ortho_height < 250 ~ "100 - 250",
    ortho_height >= 250 & ortho_height < 500 ~ "250 - 500",
    ortho_height >= 500 & ortho_height < 1000 ~ "500 - 1000",
    ortho_height >= 1000 & ortho_height < 2000 ~ "1000 - 2000",
    ortho_height >= 2000 ~ ">2000"
  )) |> 
  mutate(ortho_height_cat = factor(ortho_height_cat, 
                                   levels = c("0 - 50", "50 - 100", "100 - 250", "250 - 500", "500 - 1000", "1000 - 2000", ">2000")))


# summary of individuals by type of location data 
bbt <- bb |> 
  group_by(bid, shore_status) |>
  st_drop_geometry() |>
  summarise(
    n = n()
  )


# subset a group of bird which have lots of offshore data (for testing)

offshore <- bbt |> 
  filter(shore_status == "Offshore") |>
  filter(n > 20) |>
  select(bid)

bb_od <- bb |> 
  filter(bid %in% offshore$bid) 


```

## Individual Bird Plots 

We can explore individual birds in an interactive map. The coloured circles represent the altitude of the bird at a given location. 
You can hover over the dot to show the date or click each dot to pull up more infomation about the bird, including the date and time of the observation, the altitude, and wind speed and direction.

The black arrows with the grey bakground represent the modelled wind direction at a height of 10m above the ocean. Click on the grey marker to show the modeled wind. 

Note layers can be turned on or off using the tick boxes in the top right corner of the map. 


**Bird Argos id: 233928**

```{r pressure, echo=FALSE}
## read in compiled data with movements and limit to rufa 
xx <- unique(bb_od$bid)

bb_od <- cbind(bb_od , st_coordinates(bb_od ))

#sort(unique(round(bb_od$ortho_height,0))) # outliers

# format colours 
bb_od <- bb_od |> 
  mutate(ortho_height_cat = case_when(
      ortho_height < 50 ~ "0 - 50",
      ortho_height >= 50 & ortho_height < 100 ~ "50 - 100",
      ortho_height >= 100 & ortho_height < 250 ~ "100 - 250",
      ortho_height >= 250 & ortho_height < 500 ~ "250 - 500",
      ortho_height >= 500 & ortho_height < 1000 ~ "500 - 1000",
          ortho_height >= 1000 & ortho_height < 2000 ~ "1000 - 2000",
      ortho_height >= 2000 ~ ">2000"
    )) |> 
  mutate(ortho_height_cat = factor(ortho_height_cat, 
                                    levels = c("0 - 50", "50 - 100", "100 - 250", "250 - 500", "500 - 1000", "1000 - 2000", ">2000")))



# select an example bird 
bi <- bb_od |> filter(bid == xx[1])

# pal <- colorNumeric(
# #  #palette = "viridis",
#   palette = "magma",
#   domain = unique(bi$ortho_height))

#pal <- colorQuantile("Reds", bi$ortho_height, 4,) # pretty = FALSE)

pal <- colorFactor(
  #palette = "viridis",
  palette = "Reds",
  domain = unique(bi$ortho_height_cat))


# select an icon 

# option 1:
# shipIcon  <- makeIcon(iconUrl = fs::path("C:/r_repo/2025_REKN/REKN_altitude/Pngtreeuparrow4565998.png"),
#                       iconWidth = 20,
#                       iconHeight = 20,
#                       iconAnchorX = 20,
#                       iconAnchorY = 20
#                       )
# option 2:
# shipIcon2 <- makeAwesomeIcon(icon = shipIcon,
#                              #library = "ion",
#                              #iconColor = "black",
#                              iconRotate = bi$windangle,
#                              squareMarker = FALSE,
#                              markerColor = "lightgray")

# option 3
# used icon embedded on libraty
shipIcon3 <- awesomeIcons(
  icon = "arrow-up",
  #icon  = shipIcon,
  #library = "ion",
  markerColor = "lightgray",
  iconColor = "black",
  spin = FALSE,
  extraClasses = NULL,
  squareMarker = FALSE,
  iconRotate = bi$windangle,
  fontFamily = "monospace",
  text = NULL
)

# option 4
# shipIcon4 <- makeAwesomeIcon(icon = "arrow-up",
#                              #library = "ion",
#                              #iconColor = "black",
#                              iconRotate = bi$windangle,
#                              squareMarker = FALSE,
#                              markerColor = "lightgray")


# create map 

birdmapall <- leaflet(bi) |> 
  #addProviderTiles("CartoDB.DarkMatter") %>%
  addProviderTiles("CartoDB", group = "Background") |> 
  addPolylines(data =bi, lng = bi$X, lat = bi$Y,
               color = "grey",   opacity = 0.1, stroke = TRUE, group = "Bird trajectory") |> 
  
  addCircleMarkers(lng = bi$X, lat = bi$Y,
                   weight = 4,
                   color = ~pal(bi$ortho_height_cat),
                   fill = TRUE,
                   fillOpacity = 1,
                   label = ~bi$datetime,
                   group = "Bird Altitude",
                   radius = ~10 ,
                   popup = paste("Bird id:", bi$bid, "<br>",
                                 "Flight altitude:", round(bi$ortho_height,0), "m", "<br>",
                                 "Date:", bi$datetime,"<br>",
                                 "Wind speed:", round(bi$ws,0), "m/s", "<br>",
                                 "Wind direction:", round(bi$windangle,0), "<br>")) |>
  #addMarkers(lng = bi$X, lat = bi$Y,icon = shipIcon ) |> 
  #addPolylines(data =bi, lng = bi$X, lat = bi$Y,
  #             color = "grey",   opacity = 0.1, stroke = TRUE, group = "Bird trajectory") |> 
  addLegend("bottomright", pal = pal, values = ~bi$ortho_height_cat,
          title = "Flight Altitude (m)",
           opacity = 1) |>
   addAwesomeMarkers(
     lng = bi$X, 
     lat = bi$Y,
     icon = shipIcon3,
     popup = paste("Wind speed:", round(bi$ws,0), "m/s", "<br>"),
     group = "Wind speed and direction") |> 
  addLayersControl(
    baseGroups = 
      "Background"
     #"Positron (minimal)",
    #  "World Imagery (satellite)"
    #)
    ,
    overlayGroups = c("Bird Altitude", "Bird trajectory", "Wind speed and direction"),
    options = layersControlOptions(collapsed = FALSE)
  )


birdmapall

```


